---
output: html_document
---
# Practical Machine Learning project asignment   
Jorge Gonzalez (Feb 2015)    

##Executive Summary    
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The goal of this project is to predict the manner of performing unilateral dumbbell biceps curls based on data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The 5 possible methods include:   

  - A) exactly according to the specification   
  - B) throwing the elbows to the front  
  - C) lifting the dumbbell only halfway   
  - D) lowering the dumbbell only halfway  
  - E) throwing the hips to the front  


Libraries required.
```{r}
set.seed(90015)
library(caret)
library(corrplot)
library(kernlab)
library(knitr)
library(randomForest)

opts_chunk$set(cache = FALSE)
```


##Loading and preprocessing the data
Two csv files containing the training and test data was downloaded from Amazon's cloudfront on 08/02/2015 into a data folder in the working directory. 

```{r}
setwd("C:/COURSERA/LearningMachine/pml")
# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
url1 <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
destfile1 <- "./data/pml-training.csv"
url2 <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
destfile2 <- "./data/pml-testing.csv"

# download the file and note the time
download.file(url1, destfile = destfile1)
download.file(url2, destfile = destfile2)
dateDownloaded <- date()
```

The training data was then loaded and cleaned into R.
```{r}
# read the csv file for training 
data_training <- read.csv("./data/pml-training.csv", na.strings= c("NA",""," "))
```
There was a lot of NA values in the data which would create a lot of noise for the model, so, these columns were removed from the data set. The first eight columns that acted as identifiers for the experiment were also removed.
```{r}
# clean the data by removing columns with NAs etc
data_training_NAs <- apply(data_training, 2, function(x) {sum(is.na(x))})
data_training_clean <- data_training[,which(data_training_NAs == 0)]

# remove identifier columns such as name, timestamps etc
data_training_clean <- data_training_clean[8:length(data_training_clean)]
```


##Creating a model (Randomforest)
The cleaned training data set was split up into training and cross validation sets in a 70:30 ratio in order to train the model and then test it against data it was not specifically fitted to.
```{r}
# split the cleaned training data into training and cross validation
inTrain <- createDataPartition(y = data_training_clean$classe, p = 0.7, list = FALSE)
training <- data_training_clean[inTrain, ]
crossval <- data_training_clean[-inTrain, ]
```

A random forest model was selected to predict the classification because it has methods for balancing error in class population unbalanced data sets. The correlation between any two trees in the forest increases the forest error rate. Therefore, a correllation plot was produced in order to see how strong the variables relationships are with each other.
```{r}
# plot a correlation matrix
correlMatrix <- cor(training[, -length(training)])
corrplot(correlMatrix, order = "FPC", method = "circle", type = "lower", tl.cex = 0.6,  tl.col = rgb(0, 0, 0))
```


In this type of plot the dark red and blue colours indicate a highly negative and positive relationship respectively between the variables. AS we can see on the plot,there isn't much concern for highly correlated predictors which means that all of them can be included in the model.

Then a model was fitted with the outcome set to the training class and all the other variables used to predict.
```{r}
# fit a model to predict the classe using everything else as a predictor
model <- randomForest(classe ~ ., data = training)
model
```

The model produced a very small OOB error rate of .6%. This was deemed satisfactory enough, so we perform Cross-valiation and testing.



##Cross-validation
The model was then used to classify the remaining 30% of data. The results were placed in a confusion matrix along with the actual classifications in order to determine the accuracy of the model.
```{r}
# crossvalidate the model using the remaining 30% of data
predictCrossVal <- predict(model, crossval)
confusionMatrix(crossval$classe, predictCrossVal)
```

This model yielded a 99.5% prediction accuracy, therefore this model proved very robust and adequate to predict new data.



##Predictions
The testing data set was then loaded into R and cleaned in the same manner as before. The model was then used to predict the classifications of the 20 results of this data.
```{r}
# apply the same treatment to the final testing data
data_test <- read.csv("./data/pml-testing.csv", na.strings= c("NA",""," "))
data_test_NAs <- apply(data_test, 2, function(x) {sum(is.na(x))})
data_test_clean <- data_test[,which(data_test_NAs == 0)]
data_test_clean <- data_test_clean[8:length(data_test_clean)]

# predict the classes of the test set
predictTest <- predict(model, data_test_clean)
predictTest
```


##Conclusions

The Random Forest algorithm is well suited to generating accurate predictions for this specific dataset.
The accuracy of the model is 0.995. The expected out-of-sample error is estimated at     0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. 
Applying the random forest model to predict the 20 test cases shown above, resulted in a 100% classification rate.
